---
html_document: PA1_template.html
output:
  html_document:
    df_print: paged
title: 'Reproducible Research: Peer Assessment 1'
keep_md: yes
---

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(lattice)
```

## Loading and preprocessing the data
```{r loaddata}
rawdata <- read.csv("activity.csv")
```
  
## What is mean total number of steps taken per day?
```{r summarize, message=FALSE, warning=FALSE}
#Count the total per day
summarydata <- rawdata %>% select(date, steps) %>% 
                            group_by(date) %>% 
                            summarize(steps = sum(steps, na.rm = TRUE))

hist(summarydata$steps)

mean(summarydata$steps, na.rm = TRUE)
median(summarydata$steps, na.rm = TRUE)
```
  
## What is the average daily activity pattern?
```{r dailypattern, message=FALSE}
#Calculate the average per interval
dailydata <- rawdata %>% select(interval, steps) %>% 
                            group_by(interval) %>% 
                            summarize(steps = mean(steps, na.rm = TRUE))

plot(dailydata$interval, dailydata$steps, type = "l")

dailydata[dailydata$steps == max(dailydata$steps),]
```


## Imputing missing values
```{r missing, message=FALSE}
#Count the missing entries
sum(is.na(rawdata$steps))

#Summarydata contains the per 5min interval means for the steps value
# and given the clear effect time of day has on activity, those make good fill
# ins for missing values on a per interval basis

adapteddata <- rawdata
for(i in 1:length(adapteddata$steps)) 
{
    if(is.na(adapteddata[i,1])) 
    {
        adapteddata[i,1] <- dailydata[dailydata$interval == adapteddata[i,3],2]
    }
}

newsummarydata <- adapteddata %>% select(date, steps) %>% 
                            group_by(date) %>% 
                            summarize(steps = sum(steps, na.rm = TRUE))

hist(newsummarydata$steps)

mean(newsummarydata$steps)
median(newsummarydata$steps)
```
The mean and median are apparently both depressed by ignoring missing values and 
have increased as a result of filling them. The histogram also shows a more 
centered distribution after filling missing values.

## Are there differences in activity patterns between weekdays and weekends?
```{r weekends, message=FALSE}
#Determine which days are weekend days
adapteddata$daytype <- substr(weekdays(as.Date(adapteddata$date)), 1, 1) == 'S'
adapteddata$daytype <- as.factor(adapteddata$daytype)
levels(adapteddata$daytype) <- c("Weekday", "Weekend")

#Calculate the average per interval for each type of day 
dailydata <- adapteddata %>% select(interval, daytype, steps) %>% 
                            group_by(daytype, interval) %>% 
                            summarize(steps = mean(steps, na.rm = TRUE))
plots <- xyplot(steps ~ interval | daytype, data = dailydata, type = "l")
update(plots, layout = c(1,2))
```
  
Activity pattern does appear different from weekdays to weekend. There is a more
consistent amount of activity during daylight hours and single spike early on 
during weekdays (the slower time afterwards being the person's work or school 
day, possibly).